---
title: "Chapter 4"
author: "Icaro Franco Hernandes"
date: "2022-09-26"
output:
  pdf_document: default
  html_document: default
subtitle: 'Workflow: Basics'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE)
library(tidyverse)
library(knitr)
library(magrittr)
```



```{r}
this_is_a_long_name<-2.5
seq(1,10)

```

In the console, typing the first letter of a variable and then "ctrl+ $\uparrow$" gives you a list of all variables starting with those first letters. The `+` sign means that R is expecting more inputs 

```{r}

```


## Exercises 4.4

### 1
Why does this code not work?
```{r eval=FALSE, include=TRUE}
my_variable <- 10
my_varıable
#> Error in eval(expr, envir, enclos): object 'my_varıable' not found

```
This code does not work since the in the second line we are actually referring to a non-defined object. This happens due to a slight difference in the name of the variables: they have different "i's".

### 2
Teak each of the following R commands to that they run correctly:

```{r eval=FALSE, include=TRUE}
library(tidyverse)

ggplot(dota=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy))

fliter(mpg, cyl=8)
filter(diamond, carat>3)
```

Tweaking:

```{r, include=TRUE, eval=TRUE}
library(tidyverse)

mpg%>%
ggplot2::ggplot()+
  geom_point(aes(x=displ, y=hwy))



filter(mpg, cyl==8)
filter(ggplot2::diamonds, carat>3)
```


### 3
Press "Alt+Shift+K". What happens? How can you get to the same place using the menus?

The combination shows us a table with all the shortcuts from RStudio. The same shortcuts can be accessed in the Help tab.

