---
title: "Chapter 5"
author: "Icaro Franco Hernandes"
date: "2022-09-26"
output:
  pdf_document: default
  html_document: default
subtitle: Data Transformation
---

```{r setup, eva= TRUE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(magrittr)
library(nycflights13)
```



Remember that to view a whole data set we can execute, for example, `view(nycflights13::flights)`. This is a tibble, and tibble $\neq$ table. Tibbles work better for the tidyverse. To check what kind of variable we are working with, we can use the following command:
 
```{r}
typeof(nycflights13::flights$time_hour)
```
 
# Filtering

Selecting all flights from January first:
```{r}
nycflights13::flights%>%
  dplyr::filter(month==1,day==1)->jan1
#remeber that dplyr does not change the original dataset (always try to be as pure as possible).
```

If we want to also print the new data set, just put between parenthesis:
```{r}
(nycflights13::flights%>%
  dplyr::filter(month==1,day==1)->jan1)
```

Boolean operators are: `&` for **and**, `|` for **or** and `!` for **is not**.

Inclusion operator: `%in%`. For example:

```{r}
nycflights13::flights%>%
  dplyr::filter(month %in% c(11,12))
```
This will filter all flights that happened in november **or** december. The filter already excludes `NA` values.

## Exercises 5.2.4

### 1
Find all flights that

* Had an arrival delay of two or more hours

```{r}
nycflights13::flights%>%
  dplyr::filter(arr_delay>=120)->f1
```


* Flew to Houston (`IAH` or `HOU`)
```{r}
nycflights13::flights%>%
  dplyr::filter(dest %in% c("IAH", "HOU"))->f2
```

* Were operated by United, American, or Delta

```{r}
nycflights13::flights%>%
  dplyr::filter(carrier %in% c("UA", "AA", "DL"))->f3
```

* Departed in Summer (July, August, and September)

```{r}
nycflights13::flights%>%
  dplyr::filter(month %in% c(7, 8, 9))->fsummer
```

* Arrived more than two hours late, but did not leave late

```{r}
nycflights13::flights%>%
  dplyr::filter(arr_delay>120,dep_time<=sched_dep_time)->f5
```

* Were delayed by at least an hour, but made up over 30 minutes in flight

```{r}
nycflights13::flights%>%
  dplyr::filter(arr_delay>=60,air_time>30)->f6
```

* Departed between midnight and 6am (inclusive)

```{r}
nycflights13::flights%>%
  dplyr::filter(hour %in% c(seq(0,6)))->f7

or

nycflights13::flights%>%
  dplyr::filter(hour >= 0 & hour<= 6)->f71

# nycflights13::flights%>%
#   dplyr::filter(0 <= hour <= 6) -> this does not work!
```

### 2
Another useful `dplyr` filter helper is `between()`. What does it do? Can you use it to simplify the code needed to answer the precious questions?

According to the documentation, `between()` let us pick any values between to boundaries, and it is a shortcut for `x>= & x<=`. They would be useful in the cases where we had to filter for the summer months and the flights between midnight and 6 a.m.:

```{r}
nycflights13::flights%>%
  dplyr::filter(between(month, 7, 9))->f8

nycflights13::flights%>%
  dplyr::filter(between(hour, 0, 6))->f9
```


### 3
How many flights have a missing `dep_time`? What other variables are missing? What might these rows represent?

```{r}
nycflights13::flights%>%
  dplyr::filter(is.na(dep_time))%>%
  dplyr::summarise(n = dplyr::n())->na

na

```
Using the count operator from `dplyr` we can see that 8255 flights have missing values for the departure time. This means that theses flights were canceled. If we do not have the departure time, we also cannot check the  airtime, the departure delay and the arrival delay. **Remember this count operator (within the summarise function) from dplyr**.

###  4
Why is `NA^0`not missing? Why is `NA|TRUE` not missing? Whys is `FALSE & NA` not missing? Can you figure out the general rule? (`NA*0` is a tricky counterexample!)

```{r}
NA^0
NA|TRUE
FALSE&NA
```

Since we are working with boolean operators here, the general rule is that R avoids the NA values and does let them contaminate the operation. It is different from the case if we calculate the average of some values with an NA (in that case it does contaminate the average).

```{r}
v1<-c(1,1, NA)
mean(v1)
mean(v1, na.rm = T)

```
The command `na.rm=TRUE` discards the `NA` values from the calculation!

# Arranging



## Exercises 5.3.1

### 1
How could you use `arrange()` to sort all missing values to the start? (Hint: use `is.na()`)

```{r}
v<-tibble::tibble(
  x=(c(3,7,1,NA))
)

x=c(3,7,1,NA)

sort(x, decreasing = FALSE, na.last=FALSE)

```
I used the base R command for sorting.

### 2
Sort `flights` to find the most delayed flights. Find the flights that left earliest.

For the most delayed flights we just need to arrange in descending format:
```{r}
nycflights13::flights%>%
  dplyr::arrange(desc(dep_delay))
```

And for the earliest departures we just need to arrange them:

```{r}
nycflights13::flights%>%
  dplyr::arrange(dep_delay)

```


### 3
Sort `flights` to find the fastest (highest speed) flights

```{r}
nycflights13::flights%>%
  dplyr::select(air_time,
                dplyr::everything())%>%
  dplyr::arrange(air_time)
```


### 4
Which flights traveled the farthest? Which traveled the shortest?

The ones traveled the farthest:
```{r}
nycflights13::flights%>%
  dplyr::select(distance,
                dplyr::everything())%>%
  dplyr::arrange(desc(distance))
```

The ones that traveled the shortest:

```{r}
nycflights13::flights%>%
  dplyr::select(distance,
                dplyr::everything())%>%
  dplyr::arrange(distance)
```

# Selecting Columns

Selecting by the specific name of each column:
```{r}
nycflights13::flights%>%
  dplyr::select(
    year,
    month,
    day)
```

Selecting an interval of columns:
```{r}
nycflights13::flights%>%
  dplyr::select(
    month:arr_time
  )
```

Negative selection:
```{r}
nycflights13::flights%>%
  dplyr::select(
    -(month:arr_time))
```
The helpers are from `tidyselect`:

* `tidyselect::starts_with("")`

* `tidyselect::ends_with("")`

* `tidyselect::contains("")`

* `tidyselect::matches("")`

* `tidyselect::num_range("")`

We can select a few columns and also the rest of them:
```{r}

nycflights13::flights%>%
  dplyr::select(
    minute,
    distance,
    dplyr::everything())

```



## Exercises 5.4.1

### 1
Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from `flights`.



### 2
What happens if you include the name of a variable multiple times in a `select()` call?
```{r}
nycflights13::flights%>%
  dplyr::select(
    year,
    minute,
    year) #crime ocorre e nada acontece nesse caso

nycflights13::flights%>%
  dplyr::select(
    year,
    -(year)) #empty vector

nycflights13::flights%>%
  dplyr::select(
    (year:day),
    year)
```

If we include the same column more than one time, the selection takes into account just the first operation. Unless it is a contradictory selection, which is this case we probably get an empty vector.


### 3
What does the `any_of()` function do? Why might if be helpful in conjunction with this vector?

Lets read the documentation:
```{r}
?any_of()

vars<-c("year","month","day","dep_delay","arr_delay")

nycflights13::flights%>%
  dplyr::select(
    any_of(vars))

nycflights13::flights%>%
  dplyr::select(
    -any_of(vars))
```
The `any_of()` command takes a vector of variables and selects columns according to the variables inside the vector. The documentation states that it is useful for negative selections and also that it does not check for errors. It is not clear yet to me what are the advantages.



### 4
Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?

```{r}
nycflights13::flights%>%
  dplyr::select(contains("TIME"))

select(flights, 
       tidyselect::contains("TIME", ignore.case=FALSE))

select(flights, 
       tidyselect::contains("time", ignore.case=FALSE)) #this one should return the same results as the first.

```
The result does surprise. At a first glance we would thought that this selection would return empty vectors, however the default setting of `contains()` ignore the differences between upper and lower case. If we want to make a more precise selection (and make a distinction of upper and lower case) we can add `ignore.case=FALSE`. 

# Mutate

```{r}
nycflights13::flights%>%
  dplyr::select(
    year:day,
    tidyselect::ends_with("delay"),
    distance,
    air_time)%>%
  dplyr::mutate(
    gain=dep_delay - arr_delay,
    speed=(distance/air_time)*60)->flights_sml

```

If we want just to keep the new variables we can use `dplyr::transmute` instead.

We can use `dplyr::mutate` with any vectorised operation. If one vector is shorter than the other, the operation will be recycled (will repeat until the the end of the bigger vector).

* Integer division: `%/%`;
* Remainder: `%%`;
* Lead and lag: `lead()`, `lag()`;
* Cumulative mean: `cummean()`;


## Exercises 5.5.2

### 1
Currently `dep_time()` and `sched_dep_time` are convenient to look at, but hard to compute because they're not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight. 
```{r}
nycflights13::flights%>%
  dplyr::select(
    dep_time,
    sched_dep_time)%>%
  dplyr::mutate(
    dep_time_min=(dep_time%/%100)*60+(dep_time)%%100,
    sched_dep_time_min=(sched_dep_time%/%100)*60+(sched_dep_time)%%100)

```


### 2
Compare `air_time` with `arr_time-dep_time`. What do you expect to see? What do you see? What fo you need to do to fix it?

```{r}
nycflights13::flights%>%
  dplyr::select(
    air_time,
    arr_time,
    dep_time)%>%
  dplyr::mutate(
    diff=arr_time-dep_time)
```

The variable air time is the amount of time spent in the air, in minutes. However the arrival and departure time are in the format of HoursMinutes. So, to fix it, we must compute the difference in minutes (or convert the arrival and departure time into minutes before subtracting).

```{r}

nycflights13::flights%>%
  dplyr::select(
    air_time,
    arr_time,
    dep_time)%>%
  dplyr::mutate(
    arr_time_min = (arr_time%/%100)*60+(arr_time%%100),
    dep_time_min = (dep_time%/%100)*60+(dep_time%%100),
    air_time2 = arr_time_min-dep_time_min)

```


### 3
Compare `dep_time`, `sched_dep_time` and `dep_delay`. How would you expect those three numbers to be related?

```{r}
nycflights13::flights%>%
  dplyr::select(
    dep_time,
    sched_dep_time,
    dep_delay)
```
They are related in the following form:
```{r}
nycflights13::flights%>%
  dplyr::select(
    dep_time,
    sched_dep_time,
    dep_delay)%>%
  dplyr::mutate(
    dep_time_min = (dep_time%/%100)*60+(dep_time%%100), 
    sched_dep_time_min = (sched_dep_time%/%100)*60+(sched_dep_time%%100),
    calc=dep_time_min-sched_dep_time_min)

```


### 4
Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for `min_rank()`

```{r}

nycflights13::flights%>%
  select(
    flight,
    dep_delay)%>%
  dplyr::filter(min_rank(dep_delay)<10)

```


### 5
What does `1:3+1:10` return? Why?

```{r}
1:3+1:10
```
Since the vectors are of different length the smallest vector is repeated in the operation.

### 6
What trigonometric functions does R provide?

The basic three functions are:
* `cos(x)`    
* `sin(x)`
* `tan(x)`

These three following functions calculate the arc and the two-argument arc-tan:

* `acos(x)`
* `asin(x)`
* `atan(x)`
* `atan2(x,y)` 

The last three functions calculate the `function(pi*x)`.

* `cospi(x)`
* `sinpi(x)`
* `tanpi(x)`

# Grouped summaries with `summarise()`

```{r}
nycflights13::flights%>%
  dplyr::group_by(year, month, day)%>%
  dplyr::summarise(
    delay = mean(dep_delay, na.rm=TRUE),
    .groups = "drop")->gp
```


```{r}
nycflights13::flights%>%
  dplyr::group_by(dest)%>%
  dplyr::summarise(
    count = n(),
    dist = mean(distance, na.rm=TRUE),
    delay = mean(arr_delay, na.rm=TRUE)
  )->dest  #just to visualize the dataframe

nycflights13::flights%>%
  dplyr::group_by(dest)%>%
  dplyr::summarise(
    count = n(),
    dist = mean(distance, na.rm=TRUE),
    delay = mean(arr_delay, na.rm=TRUE))%>%
  dplyr::filter(count > 20, dest != "HNL")%>%
  ggplot2::ggplot(aes(x=dist, y=delay))+
  geom_point()+
  geom_smooth()
  
```

```{r}
nycflights13::flights%>%
  dplyr::group_by(tailnum)%>%
  summarise(
    delay = mean(arr_delay, na.rm=TRUE)
  )%>%
  ggplot2::ggplot(aes(x=delay))+
  geom_freqpoly(binwidth=10)
```


Number of flights per average delay:

```{r}
nycflights13::flights%>%
  dplyr::group_by(tailnum)%>%
  summarise(
    delay = mean(arr_delay, na.rm=TRUE),
    n=n())%>%
  dplyr::filter(n > 25)%>%
  ggplot2::ggplot(aes(x=n, y=delay))+
  geom_point(alpha = 1/10)

```


Counts and proportions of logical values: `sum(x > 10)`, `mean(y == 0)`. When used with numeric functions, `TRUE` is converted to 1 and `FALSE` to 0. This makes `sum()` and `mean()` very useful: `sum(x)` gives the number of `TRUE`s in x, and `mean(x)` gives the proportion.


## Exercises 5.6.7

### 1
Brainstorm at least 5 different ways to assess the typical delay characteristics of a group of flights:

* A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time.

* A flight is always 10 minutes late.

* A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of the time.

* 99% of the time a flight is on time. 1% of the time it's 2 hours late.

Which is more important: arrival delay or departure delay?

### 2
Come up with another approach that will give you the same output as `not_cancelled%>%count(dest)` and `not_cancelled%>%count(tailnum, wt=distance)` (without using `count()`). 

Lets check the output from the book:
```{r}
nycflights13::flights%>%
  dplyr::filter(!is.na(dep_delay), !is.na(arr_delay))->not_cancelled

not_cancelled %>% 
  count(dest)->nc

not_cancelled %>% 
  count(tailnum, wt = distance)->wtd
```


The other way to achieve the same results is grouping by and then summarizing:
```{r}
not_cancelled%>%
  dplyr::group_by(dest)%>%
  dplyr::summarise(
    n=n()
  )->my_nc

not_cancelled%>%
  dplyr::group_by(tailnum)%>%
  summarise(
    totalmiles = sum(distance)
  )->my_wtd
```


### 3
Our definition of cancelled flights (`is.na(dep_delay)|is.na(arr_dealy)`) is slightly sub optimal. Why? Which is the most important column?

This is suboptimal because the command is assessing two vectors. Instead I propose using just the variable Air Time. We can check the following code:

```{r}
nycflights13::flights%>%
  dplyr::select(
    dep_delay,
    arr_delay,
    air_time,
    dplyr::everything())%>%
  dplyr::filter(is.na(air_time))->arrg
```

Filtering for just the non-available numbers in the air time variable, we can see that is basically the same result as getting the n.a. values from departure delay **or** n.a. values from arrival delay (I believe: `is.na(air_time)==is.na(dep_delay)|is.na(arr_dealy)`)

### 4
Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?

Even if is suboptimal, I will use the book's definition of cancelled flights:

```{r}
nycflights13::flights%>%
  dplyr::filter(is.na(dep_delay)|is.na(arr_delay))%>%
  dplyr::group_by(day)%>%
  dplyr::summarise(
    avg_delay=mean(dep_delay, na.rm=TRUE),
    n=n())%>%
  ggplot2::ggplot(aes(x=n, y=avg_delay))+
  geom_point()+
  geom_smooth()


nycflights13::flights%>%
  dplyr::filter(is.na(dep_delay)|is.na(arr_delay))%>%
  dplyr::group_by(day)%>%
  dplyr::summarise(
    n=n())%>%
  ggplot2::ggplot(aes(x=day,y=n))+
  geom_point()+
  geom_smooth()
```
It does not seem to be a clear pattern.

### 5
Which carrier has the worst delays? Challenge: can you disentangle the effects of bad airport vs. bad carriers? Why/why not? (Hint: think about `flights%>%group_by(carrier,dest)%>%summarise(n())` )

```{r}
nycflights13::flights%>%
  dplyr::group_by(carrier)%>%
  summarise(
    avg_dep_delay=mean(dep_delay, na.rm=TRUE))%>%
  dplyr::arrange(desc(avg_dep_delay))

nycflights13::flights%>%
  dplyr::group_by(carrier)%>%
  summarise(
    avg_arr_delay=mean(arr_delay, na.rm=TRUE))%>%
  dplyr::arrange(desc(avg_arr_delay))
  
```
F9 (frontier) is the worst carrier for both arrival and departure delays.

Now, taking the hint from the book:

```{r}
nycflights13::flights%>%
  dplyr::group_by(carrier, dest)%>%
  summarise(n=n())->destcar

```

If we filter the flights made by frontier, we can see that they operate only between La Guardia (LGA) and Denver (DEN). Denver is one of the busiest airports in the US, it is possible that all flights related to Denver have big delays.

### 6
What does the `sort` argument do to `count()`. When might you use it?

When `sort=TRUE`, according to the documentation, the largest groups will appear at the top. It is not clear for me when this would be most useful.


```{r}
People<-tibble::tibble(
  number=c(1,2,3,4,5,6),
  group=c(1,2,1,2,1,2))

```

# Grouped Mutates (and filters)

```{r}
nycflights13::flights%>%
  dplyr::select(
    year:day,
    tidyselect::ends_with("delay"),
    distance,
    air_time)->flights_sml

flights_sml%>%
  dplyr::group_by(year, month, day)%>%
  dplyr::filter(rank(desc(arr_delay))<10)
  

```
We are taking the 10 flights with the worst delays for each day for each month

Filtering the destinations with more than a specific number of flights:

```{r}
nycflights13::flights%>%
  dplyr::group_by(dest)%>%
  dplyr::filter(n()>365)->thr
```
Here we find all **groups** that have a bigger than a specific number. For each day of each month we have a destination with more than 365 flights.



## Exercises 5.7.1

### 1
Refer back to the lists of useful mutate and filtering functions. Describe how each operation changes when you combine it with grouping.


### 2
Which plane (`tailnum`) has the worst on-time record?



### 3
What time of day should you fly if you want to avoid delays as much as possible?

For that, lets check the delays of each hour:
```{r}
nycflights13::flights%>%
  dplyr::group_by(hour)%>%
  dplyr::summarise(
    avg_dep_delay = mean(dep_delay, na.rm=TRUE),
    avg_arr_delay = mean(arr_delay, na.rm=TRUE))%>%
  ggplot2::ggplot(aes(x=hour,y=avg_dep_delay))+
  geom_smooth()+
  geom_smooth(aes(y=avg_arr_delay), color = "black")+
  geom_vline(xintercept=19,lwd=1,colour="red")
```
I am not quite sure about this, if it is in fact correct, the time with most delays is around 19:00 and with the least delays is 05:00.

### 4
For each destination, compute the total minutes of delay. For each flight, compute the proportion of the total delay for its destination.

```{r}
nycflights13::flights%>%
  dplyr::group_by(dest)%>%
  dplyr::summarise(
    total_arr_min=sum(arr_delay, na.rm = TRUE))

```



### 5
Delays are typically temporally correlated: even once the problem that caused the initial delay has been resolved, later flights are delayed to allow earlier flights to leave. Using `lag()`, explore how the delay of a flight is related to the delay of the immediately preceding flight.



### 6
Look at each destination. Can you find flights that are suspiciously fast? (i.e. flights that represent a potential data entry error). Compute the air time of a flight relative to the shortest flight to that destination. Which flights were most delayed in the air?





### 7
Find all destinations that are flown by at least two carriers. Use that information to rank the carriers.

```{r}
nycflights13::flights%>%
  dplyr::group_by(dest)%>%
  dplyr::filter(n_distinct(carrier)>=2)%>%
  dplyr::group_by(carrier)%>%
  dplyr::summarise(destinations = n_distinct(dest)) %>%
  dplyr::arrange(desc(possible_transfers))


```



### 8
For each plane, count the number of flights before the first delay of greater than 1 hour.


